---
title: "Method comparison (simulation)"
vignette: >
  %\VignetteIndexEntry{Introduction}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(fig.width=12, fig.height=7,
                      echo=TRUE, warning=FALSE, message=FALSE,
                      tidy=TRUE, collapse = TRUE, comment = "#>")
```

```{r setup}
library(panplotter)
library(tidyverse)
library(data.table)
library(ape)
library(ggthemes)
```

## Simulation

Simulate pangenome evolution with sampling error rate.


### Simulate sampling bias

```{r}
set.seed(12345)

nreps <- 5
sim_bias <- map(1:nreps, ~ {
  print(.x)
  tsim <- simulate_pan(rate = 1e-3, 
                       mean_trans_size = 1,
                       fn_error_rate=0, 
                       fp_error_rate=0,
                       ngenomes = 200)
  size <- 0
  i <- 1
  subset <- tsim
  nodes <- sample(tsim$tree$Nnode+2:tsim$tree$Nnode)
  while((size<=20) || (size>100)){
    n <- nodes[[i]]
    subset$tree <- ape::extract.clade(tsim$tree, n)
    size <- length(subset$tree$tip.label)
    i <- i+1
  }
  
  subset$pa <- subset$pa[rownames(subset$pa) %in% subset$tree$tip.label,]
  subset$pa <- subset$pa[,colSums(subset$pa)>0]
  
  return(list(full=tsim,
       subset=subset))
})

names(sim_bias) <- 1:nreps
```

Create the necessary input files and run each algorithm

#### panicmage

set up input data

```{r}
imap(sim_bias, function (rep, i) {
  imap(rep, ~{
    ape::write.tree(.x$tree, 
                  file = paste(c("./data/sampling_bias/panicmage/", nrow(.x$pa), "_bias_", .y,'_rep_',i, '.tree'), 
                               collapse = ''))
    gc <- table(factor(colSums(.x$pa), levels = 1:nrow(.x$pa)))
    writeLines(paste(gc, collapse = ' '), 
               paste(c("./data/sampling_bias/panicmage/", nrow(.x$pa), "_bias_", .y,'_rep_',i, '.txt'), 
                               collapse = ''))
  })
})
```

run panicmage

```
for f in ./data/sampling_bias/panicmage/*.tree
do
prefix="${f%.*}"
num=$(basename $f)
num="$(cut -d'_' -f1 <<<"$num")"
echo $prefix
~/Documents/panicmage/panicmage ${prefix}.tree ${prefix}.txt $num -n > ${prefix}_results.txt
done
```

load results

```{r}
resfiles <- Sys.glob("./data/sampling_bias/panicmage/*_results.txt")

panicmage_results <- map_dfr(resfiles, ~{
  print(.x)
  l <- read_lines(.x)  
  i <- which(grepl('Some characteristics.*',l))
  params <- as.numeric(gsub('.*= ', '', unlist(str_split(l[[i+1]], ' \t '))))
  tibble(
    set = gsub('_rep.*', '', gsub('.*bias_','',.x)),
    rep = gsub('_results.*', '', gsub('.*rep_','',.x)),
    theta = params[[1]],
    tho = params[[2]],
    core = params[[3]]
  )
})



pdf <- panicmage_results %>% 
  pivot_longer(cols = colnames(panicmage_results)[3:5])

ggplot(pdf, aes(x=set, y=value)) +
  geom_point() +
  facet_wrap(~name, scales = 'free_y')
```

#### Collins et al.

We first need to convert the newick files into 'tree table' format.

```
for f in ./data/sampling_bias/panicmage/*.tree
do
prefix=$(basename $f)
prefix="${prefix%.*}"
echo $prefix
perl ~/Documents/pangenome/tre2table.pl $f > ./data/sampling_bias/collins/${prefix}.txt
done
```

We can now run the scripts from Collins et al., 2012.

```{r}
source("./scripts/f-pangenome.R")

#coalescent
mymaxit <- 10000
myreltol <- 1e-6
mymodel <- "coalescent.spec" #use the coalescent tree w/G(k)
myfitting <- "chi2" #fit it using this error function
constr <- 1 # G0 constrained to the mean genome size during fitting
mymethod <- "Nelder"
    
collins_results <- imap_dfr(sim_bias, function (rep, i) {
  imap_dfr(rep, ~{
    mat <- t(.x$pa)
    ng <- nrow(.x$pa)
    G0 <- mean(colSums(mat>0)) # mean genome size measured in gene families
    f <- Sys.glob(paste(c("./data/sampling_bias/collins/*bias_", .y,'_rep_',i, '.txt'), 
                               collapse = ''))
    print(f)
    treetable <- read.table(f, sep="\t", row.names=1)
    colnames(treetable) <- c("desc_a","desc_b","dist")
    #calculate gene family frequency spectrum
    Gk <- f.getspectrum(mat)
    #set initial parameters and recursively optimize
    opt.spec.cde <- f.recurse(c(1,100),r.data=Gk, r.genomesize = G0, r.ng = ng)
    # get optimized parameters, calculate constrained parameter
    params.spec.cde <- c(opt.spec.cde$par[1],opt.spec.cde$par[1]*(G0-opt.spec.cde$par[2]),opt.spec.cde$par[2])
    #set initial parameters and recursively optimize
    pinitial.spec.c2de <- c(params.spec.cde[1]/10,params.spec.cde[2]/100,params.spec.cde[3],params.spec.cde[1]*10000)
    opt.spec.c2de <- f.recurse(pinitial.spec.c2de,r.data=Gk,r.genomesize = G0, r.ng = ng)

    # get optimized parameters, calculate constrained parameter
    params.spec.c2de <- c(opt.spec.c2de$par,opt.spec.c2de$par[4]*(
      G0-opt.spec.c2de$par[2]/opt.spec.c2de$par[1]-opt.spec.c2de$par[3]))
    print(params.spec.c2de)
    spec.c2de <- f.coalescent.spec(params.spec.c2de,ng)
    
    #summarise paramters
    return(tibble(
      set = .y,
      rep = i,
      Gess=params.spec.c2de[3],
      theta1 = params.spec.c2de[2],
      rho1 = params.spec.c2de[1],
      theta2 = params.spec.c2de[5],
      rho2 = params.spec.c2de[4],
      fslow = params.spec.c2de[2]/params.spec.c2de[1]/G0,
      ffast = params.spec.c2de[5]/params.spec.c2de[4]/G0,
      fess = params.spec.c2de[3]/G0,
      Gnew100 = f.coalescent(params.spec.c2de,100)$pan[100]-f.coalescent(params.spec.c2de,100)$pan[99],
      Gnew1000 = f.coalescent(params.spec.c2de,1000)$pan[1000]-f.coalescent(params.spec.c2de,1000)$pan[999],
      Gcore100 = f.coalescent(params.spec.c2de,100)$core[100],
      Gcore1000 = f.coalescent(params.spec.c2de,1000)$core[1000],
      Gpan100 = f.coalescent(params.spec.c2de,100)$pan[100],
      Gpan1000 = f.coalescent(params.spec.c2de,1000)$pan[1000]
    ))
  })
})

pdf <- collins_results[,1:10] %>% 
  pivot_longer(cols = colnames(collins_results)[3:10])

ggplot(pdf, aes(x=set, y=value)) +
  geom_point() +
  facet_wrap(~name, scales = 'free_y')
```


#### Finitely Many Genes model

Here we make use of the implementation of the FMG model as described in Zamani-Dahaj et al., 2016 and implemented in the Panaroo package.

```{r}
imap(sim_bias, function (rep, i) {
  imap(rep, ~{
    ape::write.tree(.x$tree, 
                  file = paste(c("./data/sampling_bias/zamani-dahaj/bias_", .y,'_rep_',i, '.tree'), 
                               collapse = ''))
    tb <- as_tibble(t(.x$pa)) %>% 
                      add_column(gene=colnames(.x$pa),  .before = 1)
    write.table(tb, sep = '\t', quote = FALSE, row.names = FALSE,
                file = paste(c("./data/sampling_bias/zamani-dahaj/bias_", .y,'_rep_',i, '_pa.txt'), 
                               collapse = ''))
  })
})
```


```
for f in ./data/sampling_bias/zamani-dahaj/*.tree
do
prefix=$(basename $f)
prefix="${prefix%.*}"
echo $prefix
python ~/Documents/panaroo/panaroo-estimate-fmg.py -o ./data/sampling_bias/zamani-dahaj/${prefix}.txt --pa ./data/sampling_bias/zamani-dahaj/${prefix}_pa.txt --tree $f
done
```

Load results

```{r}
resfiles <- Sys.glob("./data/sampling_bias/zamani-dahaj/bias_*[0-9].txt")

zamani_results <- map_dfr(resfiles, ~{
  df <- fread(.x, skip = 4, header = FALSE, 
              col.names = c('parameter','estimate','NA1','NA2'))[,1:2] %>% 
    as_tibble()%>% 
    add_column(set=gsub('_rep.*', '', gsub('.*bias_','',.x)), .before = 1) %>% 
    add_column(rep=gsub('\\.txt', '', gsub('.*rep_','',.x)), .before = 1)
  return(df)
})

pdf <- zamani_results %>% group_by(set, rep) %>%
  summarise(
    parameter=c('u','v'),
    estimate = c(estimate[which(parameter=='a')]*estimate[which(parameter=='M')],
                 estimate[which(parameter=='v')])
  )

ggplot(pdf, aes(x=set, y=estimate)) +
  geom_point()+
  facet_wrap(~parameter)
```

#### Accumulation curves (overlapping)

Accumulation curves do not use a statistical model by are very commonly used. Uncertainty is typically quantified using permutations. Here, we consider two curves to be different if there is no overlap in their curves generated by permutation.

```{r}
nperm <- 100
curves <- imap_dfr(sim_bias, function (rep, i) {
  print(i)
  return(imap_dfr(rep, ~{
    .x <- t(.x$pa)
    plotdf <- purrr::map_dfr(1:nperm, function(i){
      ppa <- .x[sample(nrow(.x), replace = FALSE), sample(ncol(.x), replace = FALSE)]
      cumlative <- rowSums(apply(ppa, 1, cumsum)>0)
      cumlative <- cumlative-cumlative[[1]]
      df <- tibble::tibble(N = 1:length(cumlative),
                           naccessory = cumlative,
                           permutation = i)
      return(df)
    }) %>%
      tibble::add_column(pangenome=.y)
  
  plotdf <- plotdf %>%
    dplyr::group_by(N,pangenome) %>%
    dplyr::summarise(
      `accessory size` = mean(naccessory),
      std = sd(naccessory)
    ) %>%
    add_column(rep=i, .before = 1) %>%
    add_column(set=.y, .before = 1)
  return(plotdf)
  }))
})

curves$group <- paste(curves$set, curves$rep, sep = '_')

ggplot(curves, aes(N, `accessory size`, col=set, fill=set, group=group)) + 
    geom_ribbon(aes(ymin = `accessory size` - std,
                    ymax = `accessory size` + std),
                alpha=0.5, col=NA) + 
    scale_color_brewer(type = 'qual', palette = 5) +
    scale_fill_brewer(type = 'qual', palette = 5) +
    geom_line(size = 1) +
    theme_bw(base_size = 14) +
    xlab("Number of genomes") +
    ylab("Accessory size") +
    labs(fill='error rate', color='error rate') 
```

#### Panstripe

```{r}
pp_results <- imap_dfr(sim_bias, function (rep, i) {
  return(imap_dfr(rep, ~{
    twdfit <- panplotter::fit_tweedie(.x$pa,.x$tree, stochastic.mapping = FALSE, nboot = 0)
    pdf <- twdfit$boot_reps %>%
      add_column(rep=i, .before = 1) %>%
      add_column(set=.y, .before = 1)
  return(pdf)
  }))
})

ggplot(pp_results, aes(x=set, y=model.core.estimate, col=rep)) +
  geom_point() 
```

####Summary plot

```{r}
# panicmage
pannic_res <- panicmage_results %>% 
  pivot_longer(cols = colnames(panicmage_results)[3:5]) %>%
  filter(name=='theta')
pannic_res$value <- c(scale(pannic_res$value))
pannic_test <- t.test(pannic_res$value[pannic_res$set=='full'],
        pannic_res$value[pannic_res$set=='subset'], 
        alternative = 'two.sided')
pannic_test

#collins
collins_res <- collins_results[,1:10] %>% 
  pivot_longer(cols = colnames(collins_results)[3:10]) %>%
  filter(name=='theta1')
collins_res$value <- c(scale(collins_res$value))
collins_test <- t.test(collins_res$value[collins_res$set=='full'],
        collins_res$value[collins_res$set=='subset'], 
        alternative = 'two.sided')
collins_test

#Zamani-Dahaj
zamani_res <- zamani_results %>%
  filter(parameter=='v')
zamani_res$estimate <- c(scale(zamani_res$estimate))
zamani_test <- t.test(zamani_res$estimate[zamani_res$set=='full'],
        zamani_res$estimate[zamani_res$set=='subset'], 
        alternative = 'two.sided')
zamani_test

#accumulation curves
curves_res <- map_dfr(split(curves %>% filter(N>1), curves$N), ~{
  .x$value = c(scale(.x$`accessory size`))
  return(.x)
})
  
curves_res <- curves_res %>% group_by(N, rep) %>%
  summarise(
    subset=value[set=='subset'],
    full=value[set=='full']
  )

curve_test <- t.test(curves_res$subset, 
       curves_res$full, 
       paired = TRUE, alternative = "two.sided")
curve_test

#panstripe
pp_results$value <- scale(pp_results$model.core.estimate)
panstripe_test <- t.test(pp_results$value[pp_results$set=='full'],
        pp_results$value[pp_results$set=='subset'], 
        alternative = 'two.sided')
panstripe_test


tests <- tibble(
  method=c('panicmage','Zamani-Dahaj','Collins','accumulation\ncurve', 'panstripe'),
  t.statistic=c(pannic_test$statistic, zamani_test$statistic, collins_test$statistic, curve_test$statistic, panstripe_test$statistic),
  p.value=c(pannic_test$p.value, zamani_test$p.value, collins_test$p.value, curve_test$p.value, panstripe_test$p.value)
)

curves_res_subset <- curves_res %>% filter(N==20)
pdf <- tibble(
  values = c(pp_results$value, pannic_res$value, zamani_res$estimate, collins_res$value, 
             curves_res_subset$subset, curves_res_subset$full),
  set = c(pp_results$set, pannic_res$set, zamani_res$set, collins_res$set,
          rep(c('full','subset'), each=nrow(curves_res_subset))),
  method = rep(c('panstripe','panicmage','Zamani-Dahaj','Collins','accumulation\ncurve'), 
               c(nrow(pp_results), nrow(pannic_res), nrow(zamani_res), nrow(collins_res), 2*nrow(curves_res_subset)))
)

pdf$method <- factor(pdf$method, levels=c('panstripe','Zamani-Dahaj','Collins','panicmage','accumulation\ncurve'))


pdf$is.sig <- ifelse(pdf$method %in% tests$method[tests$p.value<0.05],
                     'significant', 'not significant')

ggplot(pdf, aes(x=set, y=values, col=is.sig)) +
  geom_boxplot(outlier.colour = NA, position = position_dodge(width = 0.8)) +
  geom_point(size=2, position = position_dodge(width = 0.8)) +
  theme_clean(base_size = 20) +
  facet_wrap(~method, scales = 'free_x', nrow = 1) +
  scale_color_manual(values = c('#4d4d4d', '#b2182b')) +
  theme(plot.background = element_blank(),
        legend.background = element_blank(),
        axis.title.x=element_blank()) +
  # scale_alpha_discrete(guide = 'none') +
  ylab('estimated parameter value (scaled)') +
  labs(color='')

ggsave('./figures/simulation_sampling_bias_summary.png', width = 15, height = 7)
ggsave('./figures/simulation_sampling_bias_summary.pdf', width = 15, height = 7)
```
